---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(pwr4exp)
```

# Introduction

The pwr4exp R package offers functionality for power analysis and sample
size determination for some standard and common designs in animal
studies, and it can be used for various complex experimental designs.
This section provides an overview of the functionality within the
pwr4exp and explains some common arguments used in its functions. To
perform power analysis in pwr4exp involves two steps:

## Step1: creating a design object

Design objects in pwr4exp can be created using functions that generate
several standard experimental designs available in the package. These
functions include **designCRD** for complete randomized design
(**CRD**), **designRCBD** for randomized complete block design
(**RCBD**), **designLSD** for Latin Square design (**LSD**),
**designCOD** for crossover design (**COD**), and **designSPD** for
split-plot design (**SPD**). The arguments of these functions fall into
the following main categories:

1.  **Treatment structure:** The arguments `treatments`, `trt.main`, and
    `trt.sub` are used to specify the treatment structure for designs
    other than SPD, and in the case of an SPD, they refer to the main
    plot and subplot levels, respectively. The treatment structure is
    defined by an integer-valued vector, where the length of the vector
    indicates the number of treatment factors, and each value represents
    the number of levels for each factor. A maximum of two factors is
    allowed, and they are arranged in a factorial design.

    For instance, `treatments = 2` is used to define an experiment
    involving only two treatments (e.g., control vs. intervention). In
    the case of two factors, `treatments = c(2, 2)` sets up a "2x2"
    factorial design where two factors are arranged to study the main
    effects and their interactions. In the case of SPD, `trt.main` and
    `trt.sub` allow for hierarchical structuring of treatments. For
    example, `trt.main = 2` specifies two levels of a main plot
    treatment factor, and `trt.sub = c(2, 2)` defines a "2 × 2"
    factorial design at the subplot level.

2.  **Label:** The optional argument `label` accepts a list that
    specifies the names of treatment factors and their corresponding
    level labels. Each element of the list represents a factor, with the
    name of the element being the name of the factor. The value of each
    element should be a vector containing the labels for that factor's
    levels. If not provided, the default name "trt" is assigned to a
    single treatment factor, while "facA" and "facB" are used for two
    treatment factors. The levels of each factor are sequentially coded
    as "1", "2", and so on, if not specified by `label`.

    For example, `list(trt = c("ad libitum", "fasting"))` uses the
    default name for a single treatment factor, but customizes its
    levels to "ad libitum" and "fasting." For multiple factors, such as
    ``` list(feed = c("``ad libitum``", "``fasting``"), dosage = c("D0", "D1", "D2")) ```,
    the first factor is named "feed" with levels "ad libitum" and
    "fasting" and the second factor is named "dosage" with levels "D0,"
    "D1," and "D2."

3.  **Replication:** Given distinct replicating mechanisms, three
    arguments `replicates`, `blocks`, and `squares` are used to
    represent replication and indicate sample size for different
    designs. The argument `replicates` specifies the number of
    experimental units per treatment in a CRD or the number of main
    plots (i.e., the number of experimental units per treatment at main
    plot) in a SPD; `blocks` specifies the number of blocks in an RCBD;
    and `squares` specifies the number of squares in a replicated LSD or
    crossover design.

    In CRD, `replicates = 10` along with `treatments = 4` (or
    `treatments = c(2, 2)`) indicate that each treatment group consists
    of 10 experimental units, resulting in a total of 40 experimental
    units. While configuring a SPD, `replicates = 10` combined with
    `trt.main = 4` (or `trt.main = c(2, 2)`) signifies that each main
    plot treatment is replicated across 10 experimental units,
    accumulating to 40 main plots in total. In a RCBD, the argument
    `blocks = 10` used in conjunction with `treatments = 4` (or
    `treatments = c(2, 2)`) ensures that all four treatments are
    replicated across 10 different blocks, leading to a total of 40
    experimental units. For an LSD, setting `squares = 3` alongside
    `treatments = 4` (or `treatments = c(2, 2)`) implies the replication
    of the layout of a single "4×4" square 3 times, resulting a total of
    48 experimental units.

4.  **Model:** The `formula` argument specifies the formula that would
    be used to test effects in post-experimental data analysis. This
    formula follows the syntax used in R’s `lm` (procedure for linear
    models) and `lmer` (procedure for LMM) functions to specify fixed
    and random effects. Each design-generating function within the
    package comes with a default model formula. The default model
    formula incorporates interaction terms when two treatment factors
    are present and fits blocking factors as random effects.

5.  **Effect Size:** The `beta` argument represents model coefficients.
    It is noteworthy that the coefficients for categorical variables are
    the coefficients of dummy variables created using `contr.treatment`
    contrast coding in R. This coding scheme designates the first level
    of each categorical factor as the reference level and creates
    indicator variables for the other levels to compare against this
    reference. The coefficient for the reference level is the mean of
    that level. The coefficients for the other levels are interpreted as
    the difference from this reference level.

    For example, considering a factor with 2 levels (`treatments = 2`),
    control vs. intervention (optional,
    `label = list(trt = c("control", "intervention"))`), if
    `beta = c(10, 5)`, this indicates that the mean of the control group
    is 10, and the effect of an intervention is 5 units higher than the
    control.

    In another example with a "2 × 2" factorial arrangement of
    treatments, e.g, `treatments = c(2, 2)` &
    `label= list(A = c("A1", "A2"), B = c("B1", "B2"))`, assume the
    expected outcomes for each combination of the two factors are as
    follows:

    $$
    \begin{array}{c|c|c}
      & B1 & B2 \\
    \hline
    A1 & 10 & 6 \\
    A2 & 8  & 12 \\
    \end{array}
    $$

    The argument for effect size reflects as
    `beta = c(intercept = 10, A2 = -2, B2 = -4, A2B2 = 8)`, which
    indicates that the mean of the reference level (`A1B1`), the effect
    of A2 alone (`A2B1 - A1B1)`, the effect of B2 alone (`A1B2 - A1B1`),
    and the interaction between A2 and B2 (`A2B2 - A2B1 - A1B2 + A1B1`),
    i.e., the additional effect of combining A2B2 compared to what would
    be expected from the sum of individual effects of A2 and B2. It is
    important to mention that values in `beta` indicate population
    values (i.e., expectations), which are usually unknown and must be
    derived through informed assumptions. These values can either be
    directly or transformed from group means as demonstrated.

6.  **Variance-Covariance:** The `VarCov` specifies variance-covariance
    components of random effects. If there are multiple random effects
    of one grouping factor, provide variance-covariance components in a
    matrix that matches the order of the random effects specified in the
    model formula.

    For instance, the covariance matrix for random intercepts and random
    slopes for one grouping factor is structured as:

    $$
    \begin{pmatrix}
    \tau_0^2 & \tau_{12} \\
    \tau_{12} & \tau_1^2
    \end{pmatrix}
    $$

    where $\tau_0^2$ is the variance of the random intercept, $\tau_1^2$
    is the variance of the random slope, and $\tau_{12}$ is the
    covariance between them. In the standard designs available in
    pwr4exp, the corresponding LMM are typically the simplest variance
    component models, i.e., models without random slopes. For example,
    in an RCBD with block as a random effect, the input required is the
    variance between blocks $\text{VarCov} = \tau_b^2$. If there are
    multiple grouping factors, provide the variance-covariance matrix
    for each grouping factor in a list. For example, in an LSD with both
    row and column blocks as random effects, the required inputs would
    be $\text{VarCov} = list(\tau_r^2, \tau_c^2)$, representing the
    variances of the row and column blocks, respectively.

7.  **Error Variance:** `sigma2` is the variance of random error in the
    model.

A design which is not available in the package can be potentially
constructed using the function `designCustom`. The required inputs
include `design.df`, `design.name`, `formula`, `beta`, `VarCov`, and
`sigma2`. All arguments have been defined above except for `design.df`,
which indicates a data frame with columns of independent variables,
outlining the structure of the data to be collected from the experiment,
but it does not include the response variable. Optionally, the name of
the customized design can be specified using `design.name`.

## Step2: power or sample size calculation

Once the design object is created, calculating power or sample size is
straightforward. Power for omnibus tests, including main effects and
their interactions (if included in the model specified during the
creation of the design object), can be calculated using the
`pwr.anova(design = , alpha = )` function. The required inputs include:

1.  `design`, which refers to the design object,

2.  `alpha`, indicating the Type I error rate, which has a default value
    of 0.05, if not specified.

On the other hand, power for specific contrasts can be calculated using
`pwr.contrast(design = , alpha = , spec = , method = )`. The same syntax
of the emmeans package is inherited to specify contrasts of interest.
The required inputs are:

1.  `design` and `alpha` as defined in the `pwr.anova` function

2.  `spec`, which is an argument inherited from `emmeans` specifying the
    names of the factors over which the contrasts are performed, and

3.  `method`, which is another argument inherited from `emmeans`
    specifying the method of contrasts, e.g., pairwise, linear, and
    polynomials.

The minimal sample size needed to achieve a target power can be
determined using
`find_sample_size(design.quote = , alpha = , target.power = , n_init = , n_max = )`.
This procedure uses an iterative approach to find the minimum
replications by traversing through a series of integers. The required
inputs are:

1.  `design.quote`, which is a quoted design object with unknown and
    unevaluated replications to be evaluated with varying values,

2.  `alpha` as defined in the `pwr.anova` function,

3.  `target.power`, which can be a single value specifying the target
    power for all effects or a vector of values for individual target
    power levels for each effect,

4.  `n_init` is the initial replications for the iterative process, with
    the default value set to 2,

5.  `n_max` is the maximum number of replications for the iterative
    process, with the default value of 99.

This function calculates the minimum sample size necessary by
incrementally checking integers from `n_init` to `n_max`. Currently,
sample size determination is available only for omnibus tests and not
for specific contrasts in pwr4exp.

# Completely Randomized Design

## CRD with one treatment factor

The following design profile is used to create an exemplary CRD with one
treatment factor:

1.  Treatments: 1 treatment factor with 4 levels.

2.  Replicates: 8 experimental units in each of 4 groups.

3.  Mean and effect size: The mean of control group is 35, and the
    effects of other three treatments are -5, +2, and +3.

4.  Error variance: The variance of response variable is 15.

If not specified, the treatment factor is named "trt" and its levels are
labeled as "1", "2", "3", and "4" by default. The corresponding design
is generated using the `designCRD` function:

```{r}
crd1 <- designCRD(
  treatments = 4,
  replicates = 8,
  beta = c(35, -5, 2, 3),
  sigma2 = 15
)
```

Power of omnibus test (i.e., F-test) can be calculated using the
`pwr.anova` function. Under the type I error rate of 0.05, the power for
testing an overall difference among treatments is 0.95467.

```{r}
pwr.anova(design = crd1)
```

The function `pwr.contrat` can be used to assess the power for specific
contrasts. To calculate the power for detecting the differences of other
levels to the control, contrast `method = rt.vs.ctrl` for the treatment
factor (`specs = ~ trt`) can be defined according to the specification
of **emmeans** package.

```{r}
pwr.contrast(design = crd1, specs =  ~ trt, method = "trt.vs.ctrl")
```

Another example to demonstrate specifying contrast vectors involves the
contrasts of interest: trt4 vs. trt1 (control) and comparing the
combined effect of trt1 and trt2 to that of trt3 and trt4.

```{r}
pwr.contrast(design = crd1, 
             specs =  ~ trt, 
             method = list(`trt4-trt2` = c(0, -1, 0, 1),
                           `trt12-trt34` = c(.5, .5, -.5, -.5)))
```

## CRD with "2x2" factorial treatments

The following design profile is used to create an exemplary CRD with two
treatment factors:

1.  Treatments: 2 treatment factors each with 2 levels.

2.  Label: two factors are "facA" with levels "A1" and "A2", "facB" with
    levels "B1" and "B2".

3.  Replicates: 8 experimental units in each of 4 groups.

4.  Mean and effect size: The mean of the control (A1B1) is 35. The
    effect of A2 alone is an increase of 5 units, and the effect of B2
    alone is an increase of 3 units. The interaction between A2 and B2
    introduces an additional effect of -2 units, resulting the combined
    effect of A2 and B2 below (-2) the sum of their individual effects.
    The corresponding cell means are:$$
    \begin{array}{c|c|c}
      & B1 & B2 \\
    \hline
    A1 & 35 & 38 \\
    A2 & 40  & 41 \\
    \end{array}
    $$

5.  Error variance: The variance of response variable is 15.

This design can be created as follows:

```{r}
crd2 <- designCRD(
  treatments = c(2, 2),
  replicates = 8,
  label = list(facA = c("A1", "A2"), facB = c("B1", "B2")),
  beta = c(35, 5, 3, -2),
  sigma2 = 15
)
```

The power of the omnibus test:

```{r}
pwr.anova(design = crd2)
```

The power for detecting the effect of factor A, both overall and within
each level of factor B, can be assessed as follows:

```{r}
# main effect of facA
pwr.contrast(design = crd2, specs = ~ "facA", method = "pairwise")
# effect of facA under B1 and B2, respectively
pwr.contrast(design = crd2, specs = ~ facA|facB, method = "pairwise")
```

The number of units per treatment required to achieve the target power
can be determined using `find_sample_size`, which employs an iterative
approach. A quoted design object, where replicates = n remains
unevaluated, is created first.

```{r}
crd_quote <- quote(
  designCRD(
  treatments = c(2, 2),
  replicates = n,
  label = list(facA = c("A1", "A2"), facB = c("B1", "B2")),
  beta = c(35, 5, 3, -2),
  sigma2 = 15
  )
)
```

The best sample size for the target power within the range of `n_int` and `n_max` can be determined:

```{r}
find_sample_size(design.quote = crd_quote, n_init = 2, n_max = 99)
```

# Randomized Complete Block Design

# Latin Square Design

# Crossover Design

# Split-plot Design
